### 2018-10-16 Manuscript/Manuscript.tex

The proposed environmental impacts of large igneous province (LIP) eruptions are myriad (as summarized in \citealp{Ernst2017a}). One aspect of LIP emplacement that has been hypothesized to relate to long-term global climate is the effect such emplacement could have on increasing global weatherability and driving cooling. Global weatherability is the sum of factors aside from climate itself, such as the latitudinal distribution of continents and mountain belts, that contribute to overall global weathering. On a planet with high weatherability, the CO$_2$ input from volcanism can be removed via silicate weathering at a lower atmospheric CO$_2$ concentration than on a less weatherable planet. The relatively high concentrations of Ca and Mg (that ultimately sequester carbon through precipitation as carbonate) and the high weathering rates of mafic lithologies make it so that basaltic regions consume more CO$_2$ than regions where bedrock composition is closer to bulk continental crust \citep{Dessert2003a}. Data from basaltic watersheds show that chemical weathering rates are greatest in regions with high runoff, and as a result CO$_2$ consumption in basaltic regions is most pronounced in the tropical rain belt \citep{Dessert2003a, Hartmann2014a}.

Given these factors, the emplacement of LIPs in the tropics has been hypothesized to be associated with specific episodes of climatic cooling on Earth. In the Neoproterozoic Era, the emplacement of the Franklin LIP in the tropics ca. 720~Ma, in concert with elevated runoff rates associated with continental break-up, has been implicated as a major contributor to the cooling that initiated the Sturtian snowball Earth \citep{Donnadieu2004b, Cox2016a}. In the Cenozoic, the movement of the Deccan LIP into the tropical rain belt, in concert with the low-latitude emplacement of the Ethiopian Traps, has been implicated in drawing down CO$_2$ levels in the lead-up to Oligocene glaciation of Antarctica \citep{Kent2008a,Kent2013a}.

@@

After LIPs are emplaced, they will progressively erode. In order to account for this decrease in area through with time, \citet{Godderis2017a} took the approach of fitting estimates of the original surface extent and estimates of the current extent for 5 LIPs with an exponential decay function. They used the resulting exponential decay constants to develop a first-order model to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which we have estimates of the original areal extent of the province and the current areal extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate approximate model for the progressive reduction in LIP area. The best-fit exponential function results in a LIP area half-life of 29~Myr. However, since we explicitly account for LIP burial separately in our models (see below), we prefer to exclude the 6 of 19 LIPs which have been partially or completely buried in our estimation of a representative LIP area half-life, which yields a slightly higher best-fit half-life of 36~Myr (Fig. \ref{fig:LIP_preservation}). There is variability in the half-lives for LIPs, if they are each individual fit with an exponential decay function, such that they range from shorter half-lives of $\sim$20~Myr up to $\sim$120~Myr for the Kalkarindji and Deccan Traps. In the analysis of LIP area through time, we implement decay scenarios informed by this analysis. Scenario `t$_{1/2}$ = 36 Myr' uses the best-fit half-life of 36 Myr, while `t$_{1/2}$ = 120 Myr' uses the slower decay. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis. The LIP reconstruction used in this study includes pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs does not significantly impact the calculated LIP areas throughout the Phanerozoic Eon which is the focus of this analysis.

Of the tectonic factors that could alter exposed LIP area, the most consequential is near-immediate burial by sediments of LIP volcanics that are co-located with a rift basin. There are numerous examples in the record where there is partial or complete burial of a LIP associated with rifting and thermal subsidence (Table \ref{tab:LIPs}). For example, the Afar LIP is both associated with the Ethiopian Traps which form plateau flood basalts, and successful rifting in the region of the Red Sea that has resulted in burial (Fig \ref{fig:LIP_map}). To account for the rapid decrease in exposed surface area that would result from burial by sediments in a rift basin, we impose two different burial scenarios for LIPs that are co-located with rifting. The `50\% burial' scenario imposes instantaneous burial of 50$\%$ of the LIP area while the `100\% burial' imposes instantaneous burial of the entire LIP as an end-member scenario. The LIP area analysis uses all of the available combinations of these distinct decay and burial scenarios. Our LIP reconstruction differs from that of \citet{Johansson2018a} in several respects. In contrast to the decay and decay+burial scenarios implemented on estimates of original LIP extent in this study, \citet{Johansson2018a} uses a static extent for each LIP throughout the reconstruction with some of the polygons corresponding to the present-day surface extent and some representing the original extent that includes currently buried portions of the LIP (e.g. the North American Midcontinent Rift).

The original extent LIP polygons were assigned a plate ID corresponding to a tectonic unit on Earth using the polygons of \citet{Torsvik2016a} for the Phanerozoic. The LIP polygons and tectonic units were reconstructed from 520~Ma to the present utilizing the paleogeographic model of \citet{Torsvik2016a} in the spin axis reference frame (anchor plate ID of 1). This paleogeographic model was updated to include a revision to Ordovician Laurentia \citep{Swanson-Hysell2017a} and the Paleozoic of Asia \citep{Domeier2018a}. Reconstructions and area calculations within latitude bands utilized the pyGPlates function library and custom Python scripts documented within a Jupyter notebook that reproduces the analysis and development of the associated visualizations. The total LIP area and the LIP area reconstructed within the tropical rain belt (15\textdegree~S to 15\textdegree~N) were calculated for the various decay and burial scenarios at a resolution of 5~Myr (Fig. \ref{fig:LIP_area}B and C).

@@

\begin{table}[h!]
\scriptsize
\caption{Phanerozoic (and the Franklin) large igneous provinces.}
\resizebox{\textwidth}{!}{
\begin{tabular}{lc>{\raggedright}p{3.5cm}cc>{\raggedright}p{3.5cm}ccl}
\toprule
name & age & age ref. & original & present & present area ref. & present/ & half-life & buried? \\
 & (Ma) & & area & area & & original & (Myr) & \\
 & & & (Mkm$^{2}$) & (Mkm$^{2}$) & & & & \\
\midrule
Columbia River & 16 & Kasbohm and Schoene (2018) & 0.68 & 0.38 & Buchan and Ernst (2004) & 0.56 & 19.2 & no \\
Afar & 30 & Courtillot and Renne (2003) & 2.05 & 0.63 & Coffin et al. (2006) & 0.31 & 17.7 & partial \\
NAIP & 62 & Larsen et al. (2016) & 1.07 & 0.29 & Buchan and Ernst (2004), Coffin et al. (2006) & 0.27 & 33.0 & partial \\
Deccan & 66 & Schoene (2015) & 0.83 & 0.56 & Coffin et al. (2006) & 0.68 & 116.6 & no \\
Seychelles & 66 & Schoene (2015) & 0.46 & 0.00 & Coffin et al. (2006) & 0.00 & 0.0 & yes \\
Madagascar & 90 & Cucciniello et al. (2010) & 0.63 & 0.03 & Coffin et al. (2006) & 0.05 & 20.8 & no \\
Caribbean-Colombian & 94 & Loewen et al. (2013) & 0.71 & 0.13 & Coffin et al. (2006) & 0.18 & 37.6 & no \\
HALIP & 95 & Kingsbury et al. (2018) & 3.60 & 0.15 & Moosdorf and Hartmann (2015) & 0.04 & 20.7 & no \\
EQUAMP & 131 & Hollanda et al. (2016) & 0.66 & 0.01 & Hollanda et al. (2016) & 0.01 & 20.5 & no \\
Comei & 132 & Zhu et al. (2009) & 0.11 & - & - & - & - & no \\
Bunbury & 132 & Zhu et al. (2009) & 0.03 & 0.00 & Thorne et al. (2014) & 0.05 & 30.9 & no \\
Parana-Etendeka & 135 & Florisbal et al. (2014) & 3.12 & 0.40 & Coffin et al. (2016), Teixeira and Albrez (2017) & 0.13 & 45.7 & partial \\
Trap & 140 & Ernst et al. (2001) & 0.03 & 0.00 & Ernst and Buchan (2001) & 0.00 & 0.0 & no \\
NW Australia Margin & 160 & COME BACK TO THIS & 0.62 & 0.00 & Coffin et al. (2006) & 0.00 & 0.0 & yes \\
Karoo & 183 & Burgess et al. (2015) & 3.21 & 0.15 & de Kock compilation & 0.05 & 41.3 & no \\
Ferrar & 183 & Burgess et al. (2015) & 0.18 & - & - & - & - & no \\
CAMP & 201 & Blackburn et al. (2013) & 11.46 & 0.23 & Marzoli and Parisio compilation & 0.02 & 35.7 & partial \\
Siberia & 252 & Burgess and Bowring (2015) & 3.46 & 0.47 & Coffin et al. (2006) & 0.14 & 87.5 & no \\
Emeishan & 259 & Zhou et al. (2002) & 0.71 & 0.06 & Coffin et al. (2006) & 0.09 & 72.9 & no \\
Panjal-Qiangtang & 283 & Zhai et al. (2013) & 0.11 & - & - & - & - & no \\
Tarim & 290 & Xu et al. (2014) & 0.35 & - & - & - & - & no \\
Magdalen & 360 & Murphy et al. (1999) & 0.42 & - & - & - & - & no \\
Vilyui & 374 & Renne et al. (2010) & 1.14 & - & - & - & - & no \\
Kola-Dnieper & 380 & Arzamastsev and Wu (2014) & 5.90 & - & - & - & - & no \\
Suordakh & 450 & Khudoley et al. (2013) & 0.02 & - & - & - & - & no \\
Kalkarindji & 511 & Jourdan et al. (2014) & 3.54 & 0.17 & Thorne et al. (2014) & 0.05 & 116.3 & no \\
Franklin & 720 & Denyszyn et al. (2009) & 2.62 & 0.04 & Buchan and Ernst (2004) & 0.02 & 121.8 & no \\
\bottomrule
\end{tabular}
}
\label{tab:LIPs}
\end{table}

>>>>

The proposed environmental impacts of large igneous province (LIP) eruptions are myriad (as summarized in \citealp{Ernst2017a} and discussed throughout this volume). One aspect of LIP emplacement that has been hypothesized to relate to long-term global climate is the effect such emplacement could have on increasing global weatherability and driving cooling. Global weatherability is the sum of factors aside from climate itself, such as the latitudinal distribution of continents and mountain belts, that contribute to overall global weathering and associated CO$_2$ sinks. On a planet with high weatherability, the CO$_2$ input from volcanism can be removed via silicate weathering at a lower atmospheric CO$_2$ concentration than on a less weatherable planet. The relatively high concentrations of Ca and Mg (that ultimately sequester carbon through precipitation as carbonate) and the high weathering rates of mafic lithologies make it so that basaltic regions consume more CO$_2$ than regions where bedrock composition is closer to bulk continental crust \citep{Dessert2003a}. Data from basaltic watersheds show that chemical weathering rates are greatest in regions with high runoff, and as a result CO$_2$ consumption in basaltic regions is most pronounced in the tropical rain belt \citep{Dessert2003a, Hartmann2009a, Hartmann2014a}.

Given these factors, the emplacement of LIPs in the tropics has been hypothesized to be associated with specific episodes of climatic cooling on Earth. In the Neoproterozoic Era, the emplacement of the Franklin LIP in the tropics ca. 720~Ma, in concert with elevated runoff rates associated with continental break-up, has been implicated as a major contributor to the cooling that initiated the Sturtian snowball Earth \citep{Donnadieu2004b, Cox2016a}. In the Cenozoic, the movement of the Deccan LIP into the tropical rain belt, in concert with the low-latitude emplacement of the Ethiopian Traps, has been implicated in drawing down CO$_2$ levels in the lead-up to Oligocene glaciation of Antarctica \citep{Kent2008a, Kent2013a}.

@@

After LIPs are emplaced, they will progressively erode. In order to account for this decrease in area with time, \citet{Godderis2017a} took the approach of fitting estimates of the original surface extent and estimates of the current extent for 5 LIPs with an exponential decay function. They used the resulting exponential decay constants to develop a first-order model to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which we have estimates of the original areal extent of the province and the current areal extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate approximate model for the progressive reduction in LIP area. The best-fit exponential function results in a LIP area half-life of 29~Myr. However, since we explicitly account for LIP burial separately in our models (see below), we prefer to exclude the 6 of 19 LIPs which have been partially or completely buried in our estimation of a representative LIP area half-life, which yields a slightly higher best-fit half-life of 36~Myr (Fig. \ref{fig:LIP_preservation}). There is variability in the half-lives for LIPs, if they are each individual fit with an exponential decay function, such that they range from shorter half-lives of $\sim$20~Myr up to $\sim$120~Myr for the Kalkarindji and Deccan Traps. In the analysis of LIP area through time, we implement decay scenarios informed by this analysis. Scenario `t$_{1/2}$ = 36 Myr' uses the best-fit half-life of 36 Myr, while `t$_{1/2}$ = 120 Myr' uses the slower decay. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis. The LIP reconstruction used in this study includes pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs does not significantly impact the calculated LIP areas throughout the Phanerozoic Eon which is the focus of this analysis.

Of the tectonic factors that could alter exposed LIP area, the most consequential is near-immediate burial by sediments of LIP volcanics that are co-located with a rift basin. There are numerous examples in the record where there is partial or complete burial of a LIP associated with rifting and thermal subsidence (Table \ref{tab:LIPs}). For example, the Afar LIP is both associated with the Ethiopian Traps which form plateau flood basalts, and successful rifting in the region of the Red Sea that has resulted in burial (Fig \ref{fig:LIP_map}). To account for the rapid decrease in exposed surface area that would result from burial by sediments in a rift basin, we impose two different burial scenarios for LIPs that are co-located with rifting. The `50\% burial' scenario imposes instantaneous burial of 50$\%$ of the LIP area while the `100\% burial' imposes instantaneous burial of the entire LIP as an end-member scenario. The LIP area analysis uses all of the available combinations of these distinct decay and burial scenarios. Our LIP reconstruction differs from that of \citet{Johansson2018a} in several respects. In contrast to the decay and decay+burial scenarios implemented on estimates of original LIP extent in this study, \citet{Johansson2018a} uses a static extent for each LIP throughout the reconstruction with some of the polygons corresponding to the present-day surface extent and some representing the original extent that includes currently buried portions of the LIP (e.g. the Keweenawan Midcontinent Rift for the which the implemented extent is from geophysical data and largely correponds to buried subsurface exposures).

The original extent LIP polygons were assigned a plate ID corresponding to a tectonic unit on Earth using the polygons of \citet{Torsvik2016a} for the Phanerozoic. The LIP polygons and tectonic units were reconstructed from 520~Ma to the present utilizing the paleogeographic model of \citet{Torsvik2016a} in the spin axis reference frame (anchor plate ID of 1). This paleogeographic model was updated to include a revision to Ordovician Laurentia \citep{Swanson-Hysell2017a} and the Paleozoic of Asia \citep{Domeier2018a}. Reconstructions and area calculations within latitude bands utilized the pyGPlates function library and custom Python scripts documented within a Jupyter notebook that reproduces the analysis and development of the associated visualizations. The total LIP area and the LIP area reconstructed within the tropical rain belt (15\textdegree~S to 15\textdegree~N) were calculated for the various decay and burial scenarios at a resolution of 5~Myr (Fig. \ref{fig:LIP_area}B and C). We use 15\textdegree~S to 15\textdegree~N as the definition of the tropical rain belt as those latitudes approximately correspond with precipitation greater than 1.0 m/yr in modern climatalogical data \citep{Kalnay1996a}. The consistent paleolatitudes of evaporite deposits in the subtropics demonstrate the stability of the large-scale atmospheric circulation that

@@

\begin{table}[h!]
\scriptsize
\caption{Phanerozoic (and the Franklin) large igneous provinces.}
\resizebox{\textwidth}{!}{
\begin{tabular}{lc>{\raggedright}p{4cm}cc>{\raggedright}p{4cm}ccl}
\toprule
\textbf{name} & \textbf{age} & \textbf{age ref.} & \textbf{original} & \textbf{present} & \textbf{present area ref.} & \textbf{present/} & \textbf{half-life*} & \textbf{buried?} \\
 & \textbf{(Ma)} & & \textbf{area} & \textbf{area} & & \textbf{original} & \textbf{(Myr)} & \\
 & & & \textbf{(Mkm$^{2}$)} & \textbf{(Mkm$^{2}$)} & & & & \\
\midrule
Columbia River & 16 & \cite{Kasbohm2018a} & 0.68 & 0.38 & \cite{Buchan2004a} & 0.56 & 19.2 & no \\
Afar & 30 & \cite{Courtillot2003a} & 2.05 & 0.63 & \cite{Coffin2006a} & 0.31 & 17.7 & partial \\
NAIP & 62 & \cite{Larsen2015a} & 1.07 & 0.29 & \cite{Buchan2004a, Coffin2006a} & 0.27 & 33.0 & partial \\
Deccan & 66 & \cite{Schoene2014a} & 0.83 & 0.56 & \cite{Coffin2006a} & 0.68 & 116.6 & no \\
Seychelles & 66 & \cite{Schoene2014a} & 0.46 & 0.00 & \cite{Coffin2006a} & 0.00 & 0.0 & yes \\
Madagascar & 90 & \cite{Cucciniello2010a} & 0.63 & 0.03 & \cite{Coffin2006a} & 0.05 & 20.8 & no \\
Caribbean-Colombian & 94 & \cite{Loewen2013a} & 0.71 & 0.13 & \cite{Coffin2006a} & 0.18 & 37.6 & no \\
HALIP & 95 & \cite{Kingsbury2018a} & 3.60 & 0.15 & \cite{Moosdorf2015a} & 0.04 & 20.7 & no \\
EQUAMP & 131 & \cite{Hollanda2016a} & 0.66 & 0.01 & \cite{Hollanda2016a} & 0.01 & 20.5 & no \\
Comei & 132 & \cite{Zhu2009a} & 0.11 & - & - & - & - & no \\
Bunbury & 132 & \cite{Zhu2009a} & 0.03 & 0.00 & \cite{Thorne2014a} & 0.05 & 30.9 & no \\
Parana-Etendeka & 135 & \cite{Florisbal2014a} & 3.12 & 0.40 & \cite{Coffin2006a} & 0.13 & 45.7 & partial \\
Trap & 140 & \cite{Ernst2001a} & 0.03 & 0.00 & \cite{Ernst2001a} & 0.00 & 0.0 & no \\
NW Australia Margin & 160 & \cite{Pirajno2012a} & 0.62 & 0.00 & \cite{Coffin2006a} & 0.00 & 0.0 & yes \\
Karoo & 183 & \cite{Burgess2015a} & 3.21 & 0.15 & de Kock compilation & 0.05 & 41.3 & no \\
Ferrar & 183 & \cite{Burgess2015a} & 0.18 & - & - & - & - & no \\
CAMP & 201 & \cite{Blackburn2013a} & 11.46 & 0.23 & Marzoli and Parisio compilation & 0.02 & 35.7 & partial \\
Siberia & 252 & \cite{Burgess2015b} & 3.46 & 0.47 & \cite{Coffin2006a} & 0.14 & 87.5 & no \\
Emeishan & 259 & \cite{Zhou2002a} & 0.71 & 0.06 & \cite{Coffin2006a} & 0.09 & 72.9 & no \\
Panjal-Qiangtang & 283 & \cite{Zhai2013a} & 0.11 & - & - & - & - & no \\
Tarim & 290 & \cite{Xu2014a} & 0.35 & - & - & - & - & no \\
Magdalen & 360 & \cite{Brendan_Murphy1999a} & 0.42 & - & - & - & - & no \\
Vilyui & 374 & \cite{Ricci2013a} & 1.14 & - & - & - & - & no \\
Kola-Dnieper & 380 & \cite{Arzamastsev2014a} & 5.90 & - & - & - & - & no \\
Suordakh & 450 & \cite{Khudoley2013a} & 0.02 & - & - & - & - & no \\
Kalkarindji & 511 & \cite{Jourdan2014a} & 3.54 & 0.17 & \cite{Thorne2014a} & 0.05 & 116.3 & no \\
Franklin & 720 & \cite{Denyszyn2009a} & 2.62 & 0.04 & \cite{Buchan2004a} & 0.02 & 121.8 & no \\
\bottomrule
\end{tabular}
}
\begin{tablenotes}[para,flushleft]
\textbf{*}assuming exponential decay with form $N(t) = 2^{t/t_{1/2}}$
\end{tablenotes}
\label{tab:LIPs}
\end{table}
